conmat: generate synthetic contact matrices for a given age-stratified population

### 2024-08-19 paper/paper.qmd
use quarto

---
title: 'conmat: generate synthetic contact matrices for a given age-stratified population'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
- affiliation: 1,2
  name: Nick Golding
  orcid: 0000-0001-8916-5570
- affiliation: 1,3
  name: Aarathy Babu
  orcid: 
- affiliation: 4
  name: Michael Lydeamore
  orcid: 0000-0001-6515-827X
- affiliation: 1,3
  name: Chitra Saraswati
  orcid: 0000-0002-8159-0414
date: "03 May 2024"
output:
  pdf_document: default
  html_document:
    keep_md: yes
bibliography: references.bib
tags:
- epidemiology
- R
- infectious disease
affiliations:
- index: 1
  name: Telethon Kids Institute
- index: 2
  name: Curtin University
- index: 3
  name: 
- index: 4
  name: Monash University
---

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
#| comment: false
knitr::opts_chunk$set(comment = "#>",
                      echo = TRUE,
                      out.width = "95%",
                      retina = 3,
                      fig.height = 4,
                      fig.align = "center",
                      dpi = 300,
                      dev = "png")
options(tinytex.clean = FALSE)
```

```{r}
#| label: libraries
library(purrr)
library(patchwork)
library(conmat)
library(ggplot2)
```

@@

This article introduces `conmat`, an R package which generates synthetic contact matrices. Contact matrices describe the number of contacts between individuals, typically age groups. They are used to create models of infectious disease spread.

There are currently few options for a user to access synthetic contact matrices [@socialmixr; @prem2017]. Existing code to generate synthetic contact matrices from @prem2017 are not designed for replicability, and are restricted to only selected countries with no sub-national demographic estimates available.

The software exposes model fitting and prediction separately to the user. Users can fit a model based on a contact survey such as POLYMOD [@mossong2008], then predict from this model to their own data. This means users can generate synthetic contact matrices for any region, with any contact survey, provided they have the relevant data.

We demonstrate a use-case for `conmat` by creating contact matrices for a state (i.e. at the sub-national level) in Australia.

For users who do not wish to run the entire `conmat` pipeline, we have pre-generated synthetic contact matrices for 200 countries, based on a list of country names by the UN, using a model fit to the POLYMOD contact survey. These resulting synthetic contact matrices, and the associated code, can be found in the syncomat analysis pipeline ([GitHub](https://github.com/idem-lab/syncomat), [Zenodo](https://zenodo.org/records/11365943)).

@@

An prominent contact survey is the "POLYMOD" study by @mossong2008, which covered 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands, and Poland [@mossong2008].

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as contact matrices or synthetic contact matrices. A widely used approach by @prem2021 [@prem2021] produced contact matrices for 177 countries at "urban" and "rural" levels for each country.

@@

We will generate a contact matrix for Tasmania, a state in Australia, using a model fitted from the POLYMOD contact survey. We can get the age-stratified population data for Tasmania from the helper function `abs_age_state()`:

```{r}
#| label: abs-age
tasmania <- abs_age_state("TAS")
tasmania
```

We can then generate a contact matrix for Tasmania using `extrapolate_polymod()`.

@@

```{r}
#| label: autoplot-contacts
#| fig.width: 8
#| fig.height: 8
autoplot(tasmania_contact)
```

@@

1) Fit a model to predict individual contact rate, using an existing contact survey
2) Predict a contact matrix using age population data

@@

Each cell in the resulting contact matrix, indexed $i$, $j$, is the predicted number of people in age group $j$ that a single individual in age group $i$ will have contact with per day. The sum over all of the $j$ age groups for a particular age group $i$ is the predicted total number of contacts per day for each individual of age group $i$.

The six covariates are 
$|i-j|$, 
${|i-j|}^{2}$, 
$i + j$, 
$i \times j$, 
$\text{max}(i, j)$ and 
$\text{min}(i, j)$.

The six key features of the relationship between the age groups, represented by the six covariates, are displayed in the figure below. These covariates capture typical features of inter-person contact, where individuals primarily interact with people of similar age (the diagonals of the matrix), and with grandparents and/or children (the so-called 'wings' of the matrix).

```{r}
#| label: partial-plots
#| echo: FALSE
fit_home <- polymod_setting_models$home
age_grid <- create_age_grid(ages = 1:99)
term_names <- extract_term_names(fit_home)
term_var_names <- clean_term_names(term_names)
age_predictions <- predict_individual_terms(
  age_grid = age_grid,
  fit = fit_home,
  term_names = term_names,
  term_var_names = term_var_names
)

@@

plot_age_term_settings <- gg_age_terms_settings(age_predictions_all_settings)
age_predictions_long <- pivot_longer_age_preds(age_predictions)
# this is a figure we want
plot_age_predictions_long <- gg_age_partial_pred_long(age_predictions_long) +
  coord_equal() +
  labs(x = "Age from",
       y = "Age to")

age_predictions_long_sum <- add_age_partial_sum(age_predictions_long)
# this is a figure we want
plot_age_predictions_sum <- gg_age_partial_sum(age_predictions_long_sum) + coord_equal() +
  labs(x = "Age from",
       y = "Age to")
# this is a figure we want
```

```{r}
#| echo: FALSE
#| fig.height: 8
plot_all_terms_sum <- plot_age_predictions_long +
  plot_age_predictions_sum + 
  plot_layout(design = "
            A
            A
            A
            A
            A
            A
            A
            B
            ") + 
  plot_annotation(
    tag_levels = "A"
  )

@@

Note that these partial dependency plots are on the log scale.

When the six covariates are added up together for each setting (in other words, each model), then exponentiated, they show the following patterns:

```{r}
# Show combined partial dep plot (i.e. sum of the partial dependencies for all six covariates) in each setting: home, school, work and other
```

In other words, the six covariates above provide patterns that are useful in modelling the different settings,
and correspond with real-life situations of how contact would look like. 
In the home setting for example, [ #TODO describe how children interact with parents and elderly generation, grandparents ].
When the covariates for school and work are added, these covariates also provide patterns that correspond with real-life situations.
https://idem-lab.github.io/conmat/dev/articles/visualising-conmat.html
In the school setting, children tend to contact other children in the same age groups as them.
In the work setting, there are no contacts with children under the age of ten and minimal contact with adults beyond retirement age.

@@

## Strengths + Weaknesses

@@

The model structure (covariates?) provided in conmat was designed to represent the key features that are typically present in a contact matrix for different settings (work, school, home, other). Including other sources of information that may better describe these contact patterns, such as inter-generational mixing, or differences in school ages of a local demographic, may improve model performance.

@@

## Meaning of the work

@@

This work was used as a key input into several models for COVID19 in Australia and contributed directly to decisions around vaccination policy.

## Unanswered questions / Future Directions

@@

https://www.doherty.edu.au/our-work/institute-themes/viral-infectious-diseases/covid-19/covid-19-modelling/modelling

Identifying vulnerable groups and predicting how diseases spread are essential for informed public health decision-making. Infectious diseases such as influenza and coronavirus spread through human-to-human interactions, or in other words, "social contact". Quantifying social contact and its patterns can provide critical insights into how these diseases spread. [ Is this circular? ] / and how best to mitigate the spread of these diseases.

>>>>

---
title: 'conmat: generate synthetic contact matrices for a given age-stratified population'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
- affiliation: 1,2
  name: Nick Golding
  orcid: 0000-0001-8916-5570
- affiliation: 1,3
  name: Aarathy Babu
  orcid: 
- affiliation: 4
  name: Michael Lydeamore
  orcid: 0000-0001-6515-827X
- affiliation: 1,3
  name: Chitra Saraswati
  orcid: 0000-0002-8159-0414
date: today
bibliography: references.bib
tags:
- epidemiology
- R
- infectious disease
affiliations:
- index: 1
  name: Telethon Kids Institute
- index: 2
  name: Curtin University
- index: 3
  name: 
- index: 4
  name: Monash University
execute: 
  echo: true
  cache: true
format: 
  pdf: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
  html: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
options(tinytex.clean = FALSE)
```

```{r}
#| label: libraries
#| echo: false
#| output: false

library(purrr)
library(patchwork)
library(conmat)
library(ggplot2)
```

@@

Contact matrices describe the number of contacts between individuals, typically age groups. They are used to create models of infectious disease spread. `conmat` is an R package which generates synthetic contact matrices for arbitrary input demography, ready for use in infectious diseases modelling.

There are currently few options for a user to access synthetic contact matrices [@socialmixr; @prem2017]. Existing code to generate synthetic contact matrices from @prem2017 are not designed for replicability, are restricted to only selected countries, and provide no sub-national demographic estimates.

The `conmat` package exposes model fitting and prediction separately to the user. Users can fit a model based on a contact survey such as POLYMOD [@mossong2008], then predict from this model to their own demographic data. This means users can generate synthetic contact matrices for any region, with any contact survey.

We demonstrate a use-case for `conmat` by creating contact matrices for sub-national level (in this case, a state) in Australia.

For users who do not wish to run the entire `conmat` pipeline, we have pre-generated synthetic contact matrices for 200 countries, based on a list of countries from the United Nations, using a model fit to the POLYMOD contact survey. These resulting synthetic contact matrices, and the associated code, can be found in the syncomat analysis pipeline ([GitHub](https://github.com/idem-lab/syncomat), [Zenodo](https://zenodo.org/records/11365943)) [@syncomat].

@@

An prominent contact survey is the "POLYMOD" study by @mossong2008, which surveyed 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands, and Poland [@mossong2008].

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as "contact matrices" or "synthetic contact matrices". A widely used approach by @prem2021 [@prem2021] produced contact matrices for 177 countries at "urban" and "rural" levels for each country.

@@

We will generate a contact matrix for Tasmania, a state in Australia, using a model fitted from the POLYMOD contact survey. We can get the age-stratified population data for Tasmania from the Australian Bureau of Statistics (ABS) with the helper function, `abs_age_state()`:

```{r}
#| label: abs-age
tasmania <- abs_age_state("TAS")
head(tasmania)
```

We can then generate a contact matrix for Tasmania, from the POLYMOD study with `extrapolate_polymod()`.

@@

```{r}
#| label: autoplot-contacts
#| fig-width: 8
#| fig-height: 8
autoplot(tasmania_contact)
```

@@

1) Fit a model to predict individual contact rate, using an existing contact survey
2) Predict a synthetic contact matrix using age population data

@@

[njt: should we mention the link/family?]

Each cell in the resulting contact matrix, indexed ($i$, $j$), is the predicted number of people in age group $j$ that a single individual in age group $i$ will have contact with per day. The sum over all of the $j$ age groups for a particular age group $i$ is the predicted total number of contacts per day for each individual of age group $i$.

[njt: do we need to mention that we need to do the exponentiated sum, since these are on the log scale?]

The six covariates are:

- $|i-j|$, 
- ${|i-j|}^{2}$, 
- $i + j$, 
- $i \times j$, 
- $\text{max}(i, j)$ and 
- $\text{min}(i, j)$.

[njt: OK with putting these as dot points?]

These covariates capture typical features of inter-person contact, where individuals primarily interact with people of similar age (the diagonals of the matrix), and with grandparents and/or children (the so-called 'wings' of the matrix).The key features of the relationship between the age groups, represented by the six covariates, are displayed in \@ref(fig:partial-plots) for the home setting. The $|i-j|$ term gives the strong diagonal, modelling people generally living with similar age people, and the $\max(i,j)$ and $\min(i,j)$ terms give the intergenerational effect of parents and grandparents with children.

```{r}
#| label: partial-plots-create
#| echo: false

fit_home <- polymod_setting_models$home
age_grid <- create_age_grid(ages = 1:99)
term_names <- extract_term_names(fit_home)
term_var_names <- clean_term_names(term_names)
age_predictions <- predict_individual_terms(
  age_grid = age_grid,
  fit = fit_home,
  term_names = term_names,
  term_var_names = term_var_names
)

@@

plot_age_term_settings <- gg_age_terms_settings(age_predictions_all_settings)
age_predictions_long <- pivot_longer_age_preds(age_predictions)
plot_age_predictions_long <- gg_age_partial_pred_long(age_predictions_long) +
  coord_equal() +
  labs(x = "Age from",
       y = "Age to") + 
  theme(
    legend.position = "bottom"
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(x = c(0, 100), y = c(0, 100))

age_predictions_long_sum <- add_age_partial_sum(age_predictions_long)
plot_age_predictions_sum <- gg_age_partial_sum(age_predictions_long_sum) + coord_equal() +
  labs(x = "Age from",
       y = "Age to") +
  theme(
    legend.position = "bottom"
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(x = c(0, 100), y = c(0, 100))
```

```{r}
#| label: show-partial-plots
#| echo: false
#| fig-height: 8
#| fig-cap: "Partial predictive plot (A) and overall synthetic contact matrix (B) for the Poisson GAM fitted to the POLYMOD contact survey in the home setting. The strong diagonal elements, and parents/grandparents interacting with children result in the classic 'diagonal with wings' shape."
plot_all_terms_sum <- plot_age_predictions_long +
  plot_age_predictions_sum + 
  plot_layout(design = "
            AAAABBBB
            AAAABBBB
            AAAABBBB
            AAAABBBB
            ") + 
  plot_annotation(
    tag_levels = "A"
  )

@@

Visualising the partial predictive plots for other settings (school, work and other), shows patterns that correspond with real-life situations are observed. A full visualisation pipeline is available at https://idem-lab.github.io/conmat/dev/articles/visualising-conmat.html

@@

@@

The covariates used by conmat were designed to represent the key features that are typically present in a contact matrix for different settings (work, school, home, other). Including other sources of information that may better describe these contact patterns, such as inter-generational mixing, or differences in school ages of a local demographic, may improve model performance.

@@

@@

This work was used as a key input into several models for COVID-19 transmission and control in Australia and contributed to decisions around vaccination policy [@DohertyModelling].

@@
